---
title: "Data Visualization with Climate Data"
output: html_notebook
---

Data Visualization with Climate Data
Instructor: Raven Davis, raven.davis@library.gatech.edu
 

Objectives: 
1. Learn how to do data visualization using base R
2. Learn basics of ggplot2 data visualization library


Getting Started:
- You can find R at <https://cran.r-project.org/>
- You can find R Studio at <https://posit.co/>

Library resources
https://learning.oreilly.com/home/
https://library.gatech.edu/ 

The Urban Climate Lab is a group of researchers at the Georgia Institute of Technology who are exploring the connections between climate change and the built environment. Their research highlights the range of mechanisms through which land use change, both within and outside of cities, is driving ongoing warming trends and impacting human and environmental health.
https://urbanclimate.gatech.edu/
#############################################################

Loading packages

```{r}
# install.packages("tidyverse") #installs tidyverse

library(tidyverse) #loads tidyverse

```

Importing data

```{r}
getwd() #check current working directory
```

```{r}

climate_data = read_csv("clean_climate.csv")

```


Exploring the data


```{r}
glimpse(climate_data) # look at data
```



```{r}
summary(climate_data) # get summary statistics for data
```

Plotting in Base R


Scatter Plot
```{r}
plot(climate_data$Date.Sensor.ID, climate_data$X5th.Street.Bridge)
```


Line graph
```{r}
plot(climate_data$Date.Sensor.ID, climate_data$X5th.Street.Bridge, type="l") # use type ="l" to indicate that you want to create a line graph
```


Histogram
```{r}
hist(climate_data$Baker.Building)
```


Histogram with color
```{r}
hist(climate_data$Baker.Building, col = "lightblue")
```


######################################################

Plotting with ggplot


Grammar of Graphics
1. data
2. aesthetic mappings
3. geometry

```{r}
# blueprint for ggplot function
"ggplot(data = <Data>,
  mapping = aes(<Mappings>) +
  geom
"
```


Building your plot iteratively 

```{r}
ggplot(data = climate_data, 
       mapping = aes(x = Date.Sensor.ID,
                     y = avg_campus_temp,
                     color= temp_category)) +
    geom_point()
```

Data

We can use dplyr functions to extract the data we need.

```{r}
climate_data %>% 
  select(Date.Sensor.ID, avg_campus_temp, temp_category) # creates tibble from subset of data frame
```
```{r}
# uses subset of data to create plot
climate_data %>% 
  select(Date.Sensor.ID, avg_campus_temp, temp_category) %>% 
  ggplot(mapping = aes(x=Date.Sensor.ID, 
                       y=avg_campus_temp, 
                       color = temp_category)) + 
  geom_point() 
```

```{r}
# we can iteratively filter out elements from our plot
climate_data %>% 
  select(Date.Sensor.ID, avg_campus_temp, temp_category) %>% 
  filter(temp_category != "High") %>% 
  ggplot(mapping = aes(x=Date.Sensor.ID, 
                       y=avg_campus_temp, 
                       color = temp_category)) + 
  geom_point() 
```

Aesthetic mappings (many of these optional):
1. x axis
2. y axis
3. color
4. alpha
5. size
6. shape
7. fill

Axes
```{r}
# must determine our axes after we select the data we want to visualize
climate_data %>% 
  ggplot(mapping = aes(x=Date.Sensor.ID, y=Baker.Building))
```


```{r}
climate_data %>% 
  ggplot(mapping = aes(x=month, y=Baker.Building))
```

Color

For the entire plot
```{r}
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building, color="red")) +
  geom_point() # change color from inside the aesthetic function
```


```{r}
# can use color names as well as hexadecimal codes, RGB codes, and number in base R

climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building)) +
  geom_point(color="#3cb371") # change using color parameter within the geom function
```


```{r}
# can map color to a specific variable's values
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building)) +
  geom_point(aes(color=Baker.Building)) # change from within the aesthetic function within the geom function
```

```{r}
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building, color = Baker.Building)) +
  geom_point() # change from within the aesthetic function
```
```{r}
# can also set a custom color palette 
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building, color = temp_category)) +
  scale_color_manual(values=c("#56B4E9","#F0E442","#a1b382")) +
  geom_point() 
```
```{r}
# also packages of pre built color palettes
library(RColorBrewer)

climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building, color = temp_category)) +
  scale_color_brewer(palette = "Dark2") +
  geom_point()
  
```
```{r}
# install.packages("ggsci")
library("ggsci") # color palettes from some scientific journals

climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building, color = temp_category)) +
  scale_color_aaas() + #american association for the advancement of science
  geom_point()

```


Alpha -> changes the transparency of the points in your plot
```{r}
climate_data %>% 
  ggplot(mapping = aes(x=Date.Sensor.ID, y=Baker.Building)) +
  geom_point(alpha=0.8) # can try different alpha values
```

```{r}
climate_data %>% 
  ggplot(mapping = aes(x=Date.Sensor.ID, y=Baker.Building)) +
  geom_point(alpha=0.2)
```
```{r}
# can set alpha based on variable value
climate_data %>% 
  ggplot(mapping = aes(x=Date.Sensor.ID, y=Baker.Building, alpha=Baker.Building)) +
  geom_point()
```


Size

```{r}
# can change the size of all points in plot
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building, color = Baker.Building)) +
  geom_point(size=3)
```

```{r}
# can change size of points based on value of a variable
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building, color = Baker.Building, size=Baker.Building)) +
  geom_point()
```

Shape
```{r}
# can assign shape to variable inside ggplot object
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building, shape=month)) +
  geom_point()
```


```{r}
# can assign shape to variable inside geom object
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building)) +
  geom_point(aes(shape=month))
```
```{r}
#also a list of different point shapes, numbered
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building, color = Baker.Building, size=Baker.Building)) +
  geom_point(shape=21)
```
Fill

```{r}
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building, fill = Baker.Building, color = "red")) +
  geom_point(shape=21)
```
```{r}
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building)) +
  geom_point(shape=21, aes(fill = Baker.Building, color = "red")) # when mapping aesthetics to variable, use aes
```

```{r}
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building)) +
  geom_point(shape=21, fill="blue", color="red") # when you're setting the visuals of plot and not mapping visuals to a variable, don't need aes()
```

Pop Quiz!
Create a plot using the starter code below and assign color, size, and shape

```{r}
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=Baker.Building)) +
  geom_point()
```


Choosing a Geom
- Different types include scatterplots, bar charts, line graphs, histograms, density plots, boxplots, and more

```{r}
# scatterplots

climate_data %>% 
  ggplot(mapping = aes(x=Date.Sensor.ID, y=Baker.Building)) +
  geom_point()
```



```{r}
# histogram
climate_data %>% 
  ggplot(mapping = aes(x=Baker.Building)) +
  geom_histogram()
```
```{r}
climate_data %>% 
  ggplot(mapping = aes(x=Baker.Building)) +
  geom_histogram(binwidth = 1) #designate bin width value to avoid warning message
```


```{r}
#line graph
climate_data %>% 
  ggplot(aes(x=Date.Sensor.ID, y=avg_campus_temp)) +
  geom_line()
```


```{r}
climate_data %>% 
  ggplot(aes(x=month, y=Canopied.Walkway)) +
  geom_boxplot() #gives interquartile range, min and max values
```

```{r}
climate_data %>%
    ggplot(mapping = aes(x=temp_category)) +
    geom_bar()
```



Geom Positions

stacking

```{r}
# subgroups of bar chart stacked on top of each other
climate_data %>%
    ggplot(mapping = aes(x=month, fill = temp_category)) +
    geom_bar(position = "stack")
```

```{r}
climate_data %>%
    ggplot(mapping = aes(x=month, fill = temp_category)) +
    geom_bar() # can do the same thing with fill
```

```{r}
# percent stacked bar chart
climate_data %>%
    ggplot(mapping = aes(x=month, fill = temp_category)) +
    geom_bar(position = "fill")
```


dodge

```{r}
# bar chart groups are next to each other instead of stacked on top
climate_data %>%
    ggplot(mapping = aes(x=month,fill=temp_category)) +
    geom_bar(position = "dodge")
```


Jitter

```{r}
climate_data %>% 
  ggplot(aes(x=month, y=Canopied.Walkway)) +
  geom_point(alpha=0.5)
```


```{r}
climate_data %>% 
  ggplot(aes(x=month, y=Canopied.Walkway)) +
  geom_point(alpha=0.5) + 
  geom_jitter() # introduces randomness into the position of points in your plot
```


Labels and Legends

```{r}
# create title, labels for x and y axis
climate_data %>%
    ggplot(aes(x = Date.Sensor.ID, y=avg_campus_temp)) +
    geom_line(color="#ff6245") +
    labs(title = "Average Temperature Over the Summer",
         x = "Date",
         y = "Average Temperature")
```
```{r}
# add a caption
climate_data %>%
    ggplot(aes(x = Date.Sensor.ID, y=avg_campus_temp)) +
    geom_line(color="#ff6245") +
    labs(title = "Average Temperature Over the Summer",
         x = "Date",
         y = "Average Temperature",
         caption = "Source: Data collected by the Urban Climate Lab")
```
```{r}
# change label for legend
climate_data %>% 
  ggplot(aes(x=month, y=avg_campus_temp, color = month)) +
  geom_boxplot() +
   labs(title = "Average Temperature Over the Summer",
         x = "Month",
         y = "Average Temperature",
         caption = "Source: Data collected by the Urban Climate Lab",
        color = "Month") 
```
Themes

```{r}
climate_data %>% 
  ggplot(aes(x=month, y=avg_campus_temp)) +
  geom_boxplot() +
   labs(title = "Average Temperature Over the Summer",
         x = "Month",
         y = "Average Temperature",
         caption = "Source: Data collected by the Urban Climate Lab") 
  theme_bw()
```

```{r}
climate_data %>% 
  ggplot(aes(x=month, y=avg_campus_temp, color = month)) +
  geom_boxplot() +
   labs(title = "Average Temperature Over the Summer",
         x = "Month",
         y = "Average Temperature",
         caption = "Source: Data collected by the Urban Climate Lab",
        color = "Month") +
  theme_dark()
```
```{r}
# can also download theme packages
install.packages("ggthemes")
library(ggthemes)
```

```{r}
climate_data %>% 
  ggplot(aes(x=month, y=avg_campus_temp, color = month)) +
  geom_boxplot() +
   labs(title = "Average Temperature Over the Summer",
         x = "Month",
         y = "Average Temperature",
         caption = "Source: Data collected by the Urban Climate Lab",
        color = "Month") +
  theme_solarized()
```

